---
title: "Train/test QSAR workflow with gaQSAR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Train/test QSAR workflow with gaQSAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows the train/test workflow used in `gaQSAR`. The workflow is useful when a separate test set is available, or when the user wants a quick first analysis before moving to double cross-validation.

The main idea is simple. First split the compounds into a training set and a test set. Then run the genetic algorithm for several model sizes. Each run gives one `gaQSAR` object. These objects are collected in a list, so that the models can be compared by their cross-validated Q2 and their performance on the external test set.

This is the workflow:

1. prepare the QSAR data;
2. split the data into training and test compounds;
3. run `gaVariableSelection()` for several numbers of predictors;
4. predict the test compounds with `predictOOBObjects()`;
5. compare the models with `createQ2Plot()`;
6. inspect the applicability domain with `createWilliamsPlot()`;
7. select a final model and, if needed, run a permutation test.

The examples below use the `AquaticTox` data from the `QSARdata` package. For a short test run, set `smokeTest <- TRUE`.

## Load packages and helper function

```{r setup}
library(gaQSAR)
library(QSARdata)

timestampFileName <- function(fileName, time = Sys.time()) {
  paste0(format(time, "%Y%m%d_%H%M%S_"), fileName)
}
```

The timestamp helper is not needed by `gaQSAR` itself. It is only used here to save result files without overwriting earlier runs.

## Prepare the data

```{r prepare-data}
data(AquaticTox)

qsarLabel <- "AquaticTox"

qsarData <- cbind(
  activity = AquaticTox_Outcome$Activity,
  AquaticTox_moe2D[, -1],
  AquaticTox_moe3D[, -1]
)

missingColumns <- which(colSums(is.na(qsarData)) != 0)

if (length(missingColumns) > 0) {
  qsarData <- qsarData[, -missingColumns]
}
```

The first column is the response variable. The remaining columns are molecular descriptors. In this example, descriptors with missing values are removed before modelling. For other data sets, a different preprocessing step may be more appropriate.

## Choose settings

```{r settings}
smokeTest <- FALSE
permutationTest <- TRUE
nWorkers <- 1L

numVars <- 2:7
theSeeds <- 1:5

gaSettings <- list(
  pMutation = 0.2,
  pCrossover = 0.7,
  popSize = 100,
  maxIter = 300,
  interval = 50,
  elitism = 3,
  crossoverType = "gaintegerOnePointCrossover",
  KSpercentage = 0.95
)

if (smokeTest) {
  gaSettings$popSize <- 25
  gaSettings$maxIter <- 10
  gaSettings$elitism <- 2
  gaSettings$interval <- 5
  numVars <- c(3, 5, 7)
  theSeeds <- 1:2
}
```

The vector `numVars` defines the model sizes to compare. For example, `numVars <- 2:7` fits models with 2, 3, 4, 5, 6 and 7 descriptors.

The `seeds` argument is used because the genetic algorithm is stochastic. Running several seeds gives a more stable result than relying on a single GA run.

## Split the data

```{r split-data}
splitupMolecules <- splitUp(qsarData, method = "KS", pc = gaSettings$KSpercentage)

xTrain <- as.matrix(qsarData[splitupMolecules$model, -1, drop = FALSE])
yTrain <- qsarData[splitupMolecules$model, 1]

xTest <- as.matrix(qsarData[splitupMolecules$test, -1, drop = FALSE])
yTest <- qsarData[splitupMolecules$test, 1]
```

The Kennard-Stone split is used here to select a training set that covers the descriptor space. With `KSpercentage = 0.95`, the external test set is small. That can be useful for a first analysis on small data sets, but it should not be oversold as a strong external validation.

## Run the GA for several model sizes

```{r run-ga}
baseArgs <- list(
  x = xTrain,
  y = yTrain,
  popSize = gaSettings$popSize,
  pMutation = gaSettings$pMutation,
  pCrossover = gaSettings$pCrossover,
  crossoverFunc = gaSettings$crossoverType,
  elitism = gaSettings$elitism,
  maxIter = gaSettings$maxIter,
  interval = gaSettings$interval,
  seeds = theSeeds,
  verbose = TRUE
)

fitOneModelSize <- function(numberOfVariables) {
  args <- baseArgs
  args$numberOfVariables <- numberOfVariables
  do.call(gaQSAR::gaVariableSelection, args)
}

output <- lapply(numVars, fitOneModelSize)
names(output) <- paste0("p", numVars)
```

Each element of `output` is a `gaQSAR` object. The list structure is important, because the plotting and comparison functions work on a set of candidate models.

## Optional parallel execution

The same loop can be run in parallel with `future.apply`. This is useful for real analyses, but it is not shown as the default in this vignette because parallel code is less convenient for package checks and examples.

```{r parallel-run}
library(future)
library(future.apply)

useFuture <- nWorkers > 1L

if (useFuture) {
  oldPlan <- future::plan()
  on.exit(future::plan(oldPlan), add = TRUE)
  future::plan(future::multisession, workers = nWorkers)

  output <- future.apply::future_lapply(
    X = numVars,
    FUN = fitOneModelSize,
    future.seed = TRUE,
    future.packages = "gaQSAR"
  )
  names(output) <- paste0("p", numVars)
}
```

## Predict the test set

```{r predict-test}
output <- predictOOBObjects(output, xTest, yTest)
```

This adds external predictions and external validation statistics to each fitted model. The test set was not used during variable selection.

## Compare Q2 values

```{r Q2-plot}
myQ2Plot <- createQ2Plot(output, label = qsarLabel)
print(myQ2Plot)

```

The Q2 plot is used to compare model sizes. The training-set Q2 is based on LOOCV. If test set predictions were added, the external Q2 is shown as well.

For small data sets, small differences in Q2 should not be treated as decisive. A smaller model with similar Q2 may be easier to interpret and less fragile.

## Williams plot, fitness plot and Observed versus Predicted plot

```{r diverse-plot}
output <- createWilliamsPlot(
  output,
  xTrain,
  yTrain,
  xTest,
  yTest,
  residualThreshold = 2.5
)

print(plot(output[[2]]))
```

The Williams plot combines standardized residuals and leverages. It is mainly used to inspect the applicability domain and to find objects that behave differently from the rest of the data.

Objects with high leverage are not automatically wrong. They may be important compounds near the edge of the descriptor space. The plot should therefore be used as a diagnostic, not as a mechanical deletion rule.

## Select one model

```{r select-model}
Q2Values <- vapply(output, function(object) object$Q2Loocv, numeric(1))
bestIdx <- which.max(Q2Values)
bestObj <- output[[bestIdx]]

cat(sprintf("Selected model size: %d predictors\n", numVars[bestIdx]))
cat(sprintf("LOOCV Q2: %.4f\n", bestObj$Q2Loocv))

summary(bestObj)
```

This selects the model with the highest LOOCV Q2. That is a practical rule, but not the only possible rule. When several model sizes perform similarly, the smaller model is often a reasonable choice.

## Permutation test

```{r permutation-test}
if (permutationTest) {
  nPermutations <- if (smokeTest) 20 else 100

  permutationResult <- gaPermutationTest(
    bestObj,
    x = xTrain,
    y = yTrain,
    nPermutations = nPermutations,
    seed = 1,
    validateSettings = TRUE,
    verbose = TRUE,
    workers = nWorkers
  )

  print(plot(permutationResult))
  summary(permutationResult)
}
```

The permutation test scrambles the response and repeats the modelling procedure under the null situation where the descriptor matrix and the response have no real relation.

In this workflow, the permutation test is applied to the selected model size. It does not fully repeat the whole earlier choice over all values of `numVars`. That distinction matters if the permutation result is used as formal evidence.

## Save results

```{r save-results}
save(output, file = timestampFileName(paste0(qsarLabel, "Results.Rdata")))

if (exists("permutationResult")) {
  save(
    permutationResult,
    file = timestampFileName(paste0(qsarLabel, "PermutationResults.Rdata"))
  )
}
```

## Summary

The train/test workflow is useful for a direct QSAR analysis with an external test set. The result is a list of fitted `gaQSAR` objects, one for each model size. This list is then used for prediction, Q2 plotting and Williams plots.

For a less biased estimate of predictive performance, especially when model size is also chosen from the data, use the double cross-validation workflow shown in the second vignette.
